Fail Faster

### 2025-03-19 paper/paper.tex
comments on intro

Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) that
provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning to find bugs fast. \iflater\bcp{Incorporate fail fast somehow.}\fi

@@

  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
optimize functional DSLs~\cn.
We apply this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

To demonstrate the effectiveness of our approach, we apply it to \bq, a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street.
Our improved version, \name, is a direct drop-in replacement for
\bq that 
is \textit{100\% semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same sequence of values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The semantic equivalence between \name and \bq allows us to 
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

@@

  \item \textbf{Python} is home to Hypothesis, the most widely used PBT library in any language.

>>>>

Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (eDSLs) \cntxt{edsls} that
provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning to find bugs fast. \iflater\bcp{Incorporate fail fast somehow.}\fi

@@

  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
build DSLs ``without regret''~\cn.
We apply this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \todo{cite some evidence for this} \jwc{Maybe this sentence isn't needed at all.}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

To demonstrate the effectiveness of our approach, we apply it to \bq, a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street.
\jwc{Also to ScalaCheck! Let's put this up front.}
Our improved version, \name, is a direct drop-in replacement for
\bq that 
is \textit{100\% semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same sequence of values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The semantic equivalence between \name and \bq allows us to 
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

@@

  \item \textbf{Python} is home to Hypothesis, the most widely used PBT library in any language.
  \jwc{More here}
